Web Processing - Standardised GIS Analyses for Cable Route Planning

### 2023-03-12 paper/discussion.tex
grammar, style, typos
understanding

% !TeX spellcheck = en_GB
In theory, the need for computing time increases with the resolution as power of two. 
Similarly, the use of main memory increases. 
This again limits the number data points, that can be processed and resolution, and probably causes difference in increase of computation time from power of 2 to a power of 2.2, because additional slower ram moduls have been used for the aggregation of higher resolution.

@@

For b), two methods have been used to increase the similarity of the paths computed from the medium resolution raster, to the path of the highest resolution raster.
These methods are used as surrogates for the more complex calculation of the Least Cost Path with the higher resolution.
In the first method a bi-linear downsampling of the higher resolution raster has been applied and in the second method, the all touched false and true raster where averaged in different ratios, to compute the optimum weighted cost raster.
While the second method of using an averaged raster, a higher similarity to the path from the highest resolution raster, the downsampling method is simpler and does not need to be optimised for the given cost.
This disadvantage could be reduced by normalising the costs.

This early stopping may result in suboptimal paths around the end points for some edge cases, where the connection via another neighbour might be more optimal.

The set of rules that are used create the cost raster, includes a rule to create buffer around buildings which is set to the level \textit{prohibited} areas.
In all touched false rasters.
The resolution of the medium level raster needs to be high enough to show every detail.
At least in the magnitude of the minimum object size plus twice its buffer.
This is true for the 10~m resolution raster and less true for the 25~m resolution raster that misses some details for roads for all touched false raster.
Other details such as rivers and houses, are already included in the lower resolution raster, due to larger buffers.
The Least Cost Path algorithm searches for an optimal path as a line.
As lines do not have a width, the route found might contain bottlenecks, that have a smaller width than the object that should be placed there.
Therefore, the used resolution should not be smaller than, the width of the object that should be placed.
This can be avoided by downsampling, but by weighting the medium resolution all touched true or false rasters.

@@

If multiple paths are calculated from a single raster, the speed-up benefit is reduced.
Especially, pre-calculation on medium resolution raster and clipping around a buffering of the resulting medium resolution paths becomes less effective.
As the number of paths increases, fewer pixels are clipped.
The Least Cost Path algorithm does only select the single most cost-effective path.
Therefore, paths of similar, but slightly higher costs remain unknown.
In additional, slight variations on the costs rasters can lead to very different paths, although the costs will not change much.
An end-user may be interested in selecting a path from a set of similar aggregated costs and applying their own evaluation criteria.
This can be achieved a by adjusting the backtracking and return polygons, or by applying perturbation on the costs.

In this work, the intermediate cost raster layers are aggregated using the maximum function. 
Another possible aggregation function is the sum or average.
Each aggregation function can be justified, by a different interpretation of the cost and its scale.

When the \textit{prohibited} level is used as the highest level, then summing the two highest levels would result in a new highest level. 
Also, the maximum function does not interfere with the nodata value. 
Although this can be done by a nansum- / nanmean-function, if the nodata value is set to \textit{not a number} during the aggregation.
The disadvantage of aggregation with the maximum function is, that this aggregation is unable to distinguish between nuances of different overlapping intermediate costs.
On sum or average aggregated rasters, one can distinguish between, different sublevels.

The fact that all touched false rasters produce more similar results than high resolution raster, is probably due to the fact, that the default level is relatively low.
As the default level increases, the effect would probably be reduced for low resolution rasters.
For high resolution raster, the effect would still be present, because the fact, that the raster pixel centre is used for sampling, reflects the original geometry better.

This effect of the similar aggregated costs per resolution could also be seen in the test paths, even when the paths varied greatly with the change in resolution or algorithm.
This could be an indication of an even spatial distribution of the costs.

The all touched true cost raster shows every detail, but the sampling with all touched true increases the size of the features.
The fact that the aggregated costs per resolution for all touched true rasters overestimate the costs when computing the path from a low resolution raster, might be due to the fact, that the values of the costs are not uniformly distributed, but that the high costs are much more frequent, because the costs are exponential scaled.

>>>>

% !TeX spellcheck = en_GB
In theory, the need for computing time increases with the resolution as power of two. 
Similarly, the use of main memory increases. 
This again limits the number data points, that can be processed and resolution, and probably causes difference in increase of computation time from power of 2 to a power of 2.2, because additional slower ram moduls has been used for the aggregation of higher resolution.

@@

For b), two methods have been used to increase the similarity of the paths computed from the medium resolution raster, to the path of the highest resolution raster.
These methods are used as surrogates for the more complex calculation of the Least Cost Path with the higher resolution.
In the first method a bi-linear downsampling of the higher resolution raster has been applied and in the second method, the all touched set to False and True raster were averaged in different ratios, to compute the optimum weighted cost raster.
While the second method of using an averaged raster, a higher similarity to the path from the highest resolution raster, the downsampling method is simpler and does not need to be optimised for the given cost.
This disadvantage could be reduced by normalising the costs.

This shows, that the way the cost raster is created in the first place can play a crucial role, in the end result.
So that a nuance, can cause a detour.
When this behaviour occurs, the polygon may not include the Least Cost Path.
This polygon should therefore be overlapped with a polygon around the shortest path.
For the set of control paths downsampling could not outperform the original medium resolution rasters.

Early stopping may result in suboptimal paths around the end points for some edge cases, where the connection via another neighbour might be more optimal.

The set of rules, that are used to create the intermediate cost raster, includes a rule to create buffer around buildings which is set to the level \textit{prohibited} areas.
In all touched set to False rasters, the resolution of the medium level raster needs to be high enough to show every detail.
At least in the magnitude of the minimum object size plus twice its buffer.
This is true for the 10~m resolution raster and less true for the 25~m resolution raster, that misses some details for roads for all touched false raster.
Other details such as rivers and houses, are already included in the lower resolution raster, due to larger buffers.
The raster with all~touched set to False might miss some objects, but the missing chance is propotional to the object size and the extent of overestimation is reduced.
The Least Cost Path algorithm searches for an optimal path as a line.
As lines do not have a width, the route found might contain bottlenecks, that have a smaller width than the object that should be placed there.
Therefore, the used resolution should not be smaller than, the width of the object that should be placed.
This can not be avoided by downsampling, but by weighting the medium resolution all~touched True and False rasters.

@@

If multiple paths are calculated from a single raster, the speed-up benefit is reduced.
Especially, pre-calculation on medium resolution raster and clipping around a buffering of the resulting medium resolution paths becomes less effective, because as the number of paths increases, fewer pixels are clipped.
The Least Cost Path algorithm does only select the single most cost-effective path.
Therefore, paths of similar, but slightly higher costs remain unknown.
In additional, slight variations on the costs rasters can lead to very different paths, although the costs will not change much.
An end-user may be interested in selecting a path from a set of similar aggregated costs and applying their own evaluation criteria.
This can be achieved a by adjusting the backtracking and return polygons, or by applying perturbation on the costs.

In this work, the intermediate cost raster layer are aggregated using the maximum function. 
Other possible aggregation functions are the sum and average functions.
Each aggregation function can be justified, by a different interpretation of the cost and its scale.

When the \textit{Prohibited} level is used as the highest level, then summing the two highest levels would result in a new highest level. 
Also, the maximum function does not interfere with the nodata value. 
This can be done by a nansum- / nanmean-function, if the nodata value is set to \textit{not a number} during the aggregation.
The disadvantage of aggregation with the maximum function is, that this aggregation is unable to distinguish between nuances of different overlapping intermediate costs.
On sum or average aggregated rasters, one can distinguish between, different sublevels.

The fact that all touched false rasters produce more similar results than high resolution raster, is probably due to the fact, that the default level is relatively low.
As the default level increases, the effect would probably be reduced for low resolution rasters.
For low resolution raster, the effect would still be present, because the fact, that the raster pixel centre is used for sampling, reflects the original geometry better.

This effect of the similar aggregated costs per resolution could also be seen in the test paths, even when the paths varied greatly.
This could be an indication of an even spatial distribution of the costs.

The all~touched set to True cost raster shows every detail, but the sampling with all~touched set to True increases the size of the features.
The fact that the aggregated costs per resolution for all touched true rasters overestimate the costs when computing the path from a low resolution raster, might be due to the fact, that the high costs are much more frequent, because the costs are exponential scaled.

### 2023-03-12 paper/methods.tex
grammar, style, typos
understanding

The least cost path algorithm is used to plan a possible route for a power line.
The least cost path algorithm is a Dijkstra algorithm~\cite{dijkstra_note_1959} applied on a raster map.
The vertices of the graph are the pixel centres, that are connected to the eight neighbouring pixel centres via the edges.
This makes it possible to find routes on graphs that are not predefined, such as road networks.
The weights of the edges are the local costs of travel from one pixel center to the neighbouring pixel center.
The costs can be physical costs, such as the local slope, but also can be composed of other factors as the acceptance rate to transverse a given land usage.
The Least Cost Path algorithm consists of at least two successive steps.
1) The first step is to aggregate the costs of travelling from the starting point to a given set of end points.
This step generates the aggregated cost raster of travelling from the starting point to any point of the cost raster.
2) In the next step the backtracking, the route of the actual Least Cost Path is calculated.
For each end point the path via the lowest cost neighbour is taken until the starting point is reached.

Some implementations switch the role of start and ending points, so that either many start points and one single end point, or one single end point and many end points can be used.
In some implementation there is an extra step between 1) and 2) that calculates cost-weight direction raster, that encodes the direction of the shortest
path to the starting point as integer values.

@@

PyWPS~\cite{noauthor_welcome_2016} is used to provide the least cost path algorithm as a \acrfull{wps} in combination with flask~\cite{noauthor_flask_nodate}.
As client Birdy~\cite{noauthor_birdy_nodate} connects to the \acrshort{wps}, sends the cost raster and receives the resulting least cost path.
The initial implementation of the Least Cost Path algorithm is based on the implementation for the QGIS-Plugin
'Least Cost Path'~\cite{noauthor_leastcostpathdijkstra_algorithmpy_2022} in version 1.0, but refactored to optionally export the aggregated costs.

To computed intermediate cost cost raster the different layers of the different entities are optionally filtered, buffered and then rasterised.
Filtering the layers of the vector files for special attributes enables further differentiation.
For examples, it is possible to differentiate between heath and uncultivated land in the land cover.
Adding a buffer can be used either to convert a line object such as power lines and roads into a polygon (with the
correct physical width), or to add minimum distance from an existing of planed area to the new power line.
Each of theses intermediate rasters are given a weight (cost) which expresses the cost of using land covered by this layer.
In the final cost raster costs of all intermediate rasters are aggregated with the maximum function.
Thus, an area covered by several layers is uniformly associated with the highest cost.
Any place in the study area, that is not covered by any layer and thus does not yet have a weight, is given the default cost.

The costs have been grouped into five different levels~(see table~\ref{tab:1}), starting from \textit{Preferential} areas with
very low costs, via \textit{No restriction}s, which is the default, used when no other layers are covering the local area,
to \textit{Restricted}, \textit{Strongly Restricted} and \textit{Prohibited} areas with high costs.
These higher costs resemble the degree to which a place with this cost should be avoided, while routing the path.
The ratio of the higher costs to the lower costs directly translates into the additional detour in pixels
the algorithm is willing to go.
Thus, as \textit{prohibited} areas describe a legal obligation, not to use these areas or only to the utmost minimum,
the weight that resembles the costs for these types of areas, is set especially high.\\

All these layers are provided as vectors.
The Least Cost Path algorithm uses raster data.
Rasterisation transforms a vector into a raster.
Rasterisation can be done in two different ways.
In both ways, the rasterisation can be imaged, as the old vector is superimposed on the new raster grid with the new
given resolution and the new affine transformation and the coordinate reference system of the vector.
Both rasterisation techniques differ in the selection of the pixel, that describe the original polygon.
The pixels can be selected, if either the centre of that pixel is overlapped
by the geo-object, or any part of the pixel is overlaid.
Setting all touched to True, implies the version with any part of the pixel selected.
The version where an overlapped pixel centre is require is setting the parameter all touched to False.
All touched set to False is considered the default (see figure~\ref{fig:alltouched}).
\begin{figure}[!ht]
	\centering

@@

			\draw[step=1cm,gray,very thin] (9.1,-1.9) grid (16.9,5.9);
			\draw[fill=blue] (10,0)--(10,5)--(11,5)--(11,4)--(15,4)--(15,1)--(13,1)--(13,0)--cycle;
		\end{tikzpicture}%
	}%
	\caption{Graphical example for rasterise a vector (left blue), to a raster (right blue) with either all touched set to False (above), or True (below).}
	\label{fig:alltouched}

@@

The complete list of layers and the applied processing steps, can be found in \textbf{Supplement S1}.

All three steps of the generation of the least cost path: generation of the cost raster, aggregation and backtracking is shown with an example for a cost raster of 50~m resolution and all touched set to False (see \ref{fig:costs2path}).

The chosen implementation applies early stopping.
Therefore, the costs for points that are not needed to try to connect to the end point are not aggregated (see figure \ref{fig:aggregation}).
After finding an aggregated cost for every end point, the aggregation stops and the backtracking starts.
Because the path ends at a power transformer, which is a building type, the paths ends at in a \textit{Prohibited} area.
Therefore, areas even further away from the starting point have been explored first.

@@

	\caption{Figures of the cost raster and the resulting aggregated costs and the Least Cost Path for a resolution of 50~m, all touched set to False.}
	\label{fig:costs2path}
\end{figure}

@@

For low resolution rasterisation with all touched set to True, will show every detail, but the objects are enlarged.
When all touched is to False the object only appears, when the is situated at the pixel centre.
Thus, this might be used as surrogate, that expresses the likelihood of the object to be sampled and correlates with the object size  compared to the pixel size.
At high resolution the set all touched to True still overestimates the object size, but the extent is limited.
Setting all touched to False will include all objects for high resolution.
This setting is most realistic, because the over- and underestimation of the object size is limited to half a pixel size in every direction.
The best method should to use the percentage of the pixel coverage by the object as weight, which is not possible.
As an alternative switching between setting all touched True and False may result in a better assessment of the true costs.
When superimposing the resulting costs rasters these map will include both aspects of the correctness: showing every detail and statistically distribute better the real cost.
Another method to achieve the same is to downsample high resolution raster.

>>>>

The Least Cost Path algorithm is used to plan a possible route for a power line.
The Least Cost Path algorithm is a Dijkstra algorithm~\cite{dijkstra_note_1959} applied on a raster map.
The vertices of the graph are the pixel centres, that are connected to the eight neighbouring pixel centres via the edges.
This makes it possible to find routes on graphs that are not predefined, such as road networks.
The weights of the edges are the local costs of transit from one pixel center to the neighbouring pixel center.
The costs can be physical costs, such as the local slope, but also can be composed of other factors as the acceptance rate to transverse a given land usage.
The Least Cost Path algorithm consists of at least two successive steps.
1) The first step is to aggregate the costs of travelling from the starting point to a given set of end points.
This step generates the aggregated cost raster of travelling from the starting point to any point of the cost raster.
2) In the next step the backtracking, the route of the actual Least Cost Path is calculated.
For each end point the path via the lowest cost neighbour is taken until the starting point is reached.

Some implementations switch the roles of start and ending points, so that either many start points and one single end point, or one single end point and many end points can be used.
In some implementations there is an extra step between 1) and 2) that calculates cost-weight direction raster, that encodes the direction of the shortest
path to the starting point as integer values.

@@

PyWPS~\cite{noauthor_welcome_2016} is used to provide the Least Cost Path algorithm as a \acrfull{wps} in combination with flask~\cite{noauthor_flask_nodate}.
As client Birdy~\cite{noauthor_birdy_nodate} connects to the \acrshort{wps}, sends the cost raster and receives the resulting Least Cost Path.
The initial implementation of the Least Cost Path algorithm is based on the implementation for the QGIS-Plugin
'Least Cost Path'~\cite{noauthor_leastcostpathdijkstra_algorithmpy_2022} in version 1.0, but refactored to optionally export the aggregated costs.

In order to compute intermediate cost raster the different vector layers of the different entities are optionally filtered, buffered and then rasterised.
Filtering the layers of the vector files for special attributes enables further differentiation.
For example, it is possible to differentiate between heath and uncultivated land in the land cover.
Adding a buffer can be used either to convert a line object such as a power line and a road into a polygon (with the
correct physical width), or to add minimum distance from an existing of planed area to the new power line.
Each of theses intermediate rasters are given a weight (cost) which expresses the cost of using land covered by this layer.
In the final cost raster costs of all intermediate rasters are aggregated with the maximum function.
Thus, an area covered by several layers is uniformly associated with the highest cost.
Any place in the study area, that is not covered by any layer and thus does not yet have a weight, is given the default cost.

The costs have been grouped into five different levels~(see table~\ref{tab:1}), starting from \textit{Preferential} areas with
very low costs, via \textit{No restriction}s, which is the default, used when no other layers are covering the local area,
to \textit{Restricted}, \textit{Strongly Restricted} and \textit{Prohibited} areas with high costs.
These higher costs represent the degree to which a place with this cost should be avoided, while routing the path.
The ratio of the higher costs to the lower costs equals the detour
the algorithm is willing to go.
Thus, as \textit{prohibited} areas describe a legal obligation, not to use these areas or only to the utmost minimum,
the weight that resembles the costs for these types of areas is set especially high.\\

All these layers are provided as vectors.
The Least Cost Path algorithm uses raster data.
Rasterisation transforms a vector into a raster.
Rasterisation can be done in two different ways.
In both ways, the rasterisation can be imaged, as the old vector is superimposed on the new raster grid with the new
given resolution and the new affine transformation and the coordinate reference system of the vector.
Both rasterisation techniques differ in the selection of the pixel, that describe the original polygon.
A pixels will be selected, if either the centre of that pixel is overlapped
by the geo-object, or any part of the pixel is overlaid.
Setting all~touched to True implies the version with any part of the pixel selected.
The version, where an overlapped pixel centre is required, is setting the parameter all~touched to False.
All~touched set to False is considered the default (see figure~\ref{fig:alltouched}).
\begin{figure}[!ht]
	\centering

@@

			\draw[step=1cm,gray,very thin] (9.1,-1.9) grid (16.9,5.9);
			\draw[fill=blue] (10,0)--(10,5)--(11,5)--(11,4)--(15,4)--(15,1)--(13,1)--(13,0)--cycle;
		\end{tikzpicture}%
	}%
	\caption{Graphical example the rasterisation of a vector (left blue), to a raster (right blue) with either all~touched set to False (above), or True (below).}
	\label{fig:alltouched}

@@

The complete list of layers and the applied processing steps can be found in \textbf{Supplement S1}.

All three steps of the generation of the Least Cost Path: generation of the cost raster, aggregation and backtracking is shown with an example for a cost raster of 50~m resolution and all~touched set to False (see \ref{fig:costs2path}).

The chosen implementation applies early stopping.
Therefore, the costs for points that are not needed to try to connect to the end point are not aggregated (see figure \ref{fig:aggregation}).
After finding an aggregated cost for every end point, the aggregation stops and the backtracking starts.
Because the path ends at a power transformer, which is a building type, the paths end at in a \textit{Prohibited} area.
Therefore, areas even further away from the starting point have been explored first.

@@

	\caption{Figures of the cost raster and the resulting aggregated costs and the Least Cost Path for a resolution of 50~m, all~touched set to False.}
	\label{fig:costs2path}
\end{figure}

@@

For low resolution rasterisation, with all~touched set to True will show every detail, but the objects are enlarged.
When all~touched is to False the object only appears, when the is situated at the pixel centre.
Thus, this might be used as surrogate, that expresses the likelihood of the object to be sampled and correlates with the object size  compared to the pixel size.
At high resolution the set all~touched to True still overestimates the object size, but the extent is limited.
Setting all~touched to False will include all objects for high resolution.
This setting is most realistic, because the over- and underestimation of the object size is limited to half a pixel size in every direction.
The best method should be to use the percentage of the pixel coverage by the object as weight, which is not possible.
As an alternative, switching between setting all~touched True and False may result in a better assessment of the true costs.
When superimposing the resulting cost raster, these map will include both aspects of the correctness: showing every detail and statistically distribute better the real cost.
Another method to achieve the same is to downsample high resolution raster.

### 2023-03-12 paper/paper-CableRoute.tex
grammar, style, typos
understanding

	\begin{abstract}
		Least Cost Paths for a power line from a power station at the cost to a power station further inland are calculated.
		Downsampling and clipping are investigated as techniques to speed up the calculation.
		The paths from the clipped cost raster often are the same as the original paths.
		The speed-up of the clipping can not be known in advance.
	\end{abstract}

@@

	%%
	%% Keywords. The author(s) should pick words that accurately describe
	%% the work being presented. Separate the keywords with commas.
	\keywords{GIS, least cost path}

@@

	However, it is not enough to find the shortest route when building new power lines.
	Other parameters have to be taken into consideration.
	As the steepness of a road or the soil, play an important role for the building cost of a road or pipeline~\cite{suleiman_optimal_2015}.
	When planing the additional routes for a power grid, further aspects such as legal regulations and acceptance
	by the local population have to be taken in consideration.
	Also technical aspects, as the effects on the grid stability are further points to take into consideration~\cite{schafer_understanding_2022}.

	Due to the increasing demand for renewable energy wind offshore wind turbines supply 5.5~\% annual percentage in 2020 of the German energy mix~\cite{noauthor_nettostromerzeugung_2021}.

	The different models the power systems, that are used in the expert system range from not modelling the grid at all to models that include Kirchhoffs laws~\cite{bertsch_participatory_2016}.
	The calculation of a power system is complex, because changing one edge changes the flow in all parts of grid and the actual grid data are confidential~\cite{bertsch_participatory_2016}.
	Some models not only consider the grid of one nation, but consider neighbouring states or the whole European power system~\cite{dietrich_will_2010}.	
	Older power grid models are e.g.\ optimising the well fare, but economic modelling is not sufficient for planning modern routes, as the aspect of environmental sustainability, security of supply and the public acceptance play an increasingly important role~\cite{hauff_gesellschaftliche_2011} in modern planning.

@@

	The uncertainties associated with \acrshort{MCDA} data uncertainties, preferential uncertainties and model uncertainties are investigated by a with a sensitivity analysis~\cite{bertsch_participatory_2016} or simulation~\cite{butler_simulation_1997}.
	Both inter- and intra-criteria preferential uncertainties~\cite{bertsch_sensitivity_2007} can be considered.

@@

	\section{Conclusion}\label{sec:conclusion}
	The cost of computing of the Least Cost Path scales with the square of the resolution.
	The difference between the aggregated costs of the paths per resolution computed from the all touched false raster and the all touched true raster only shows a linear decrease.
	Therefore, the gain in accuracy per compute time decreases.
	The presented methods have attempted to circumvent this bottleneck.
	When these strategies are applied to a medium resolution raster, the compute time for high resolution results can successfully be reduced, without compromising the run time for worse distances to the least cost path from the high resolution raster.
	The paths can change significantly even for a small change in total cost.
	Therefore, to develop an alternative backtracking algorithm, that generates a corridor of costs might be a good strategy, to offer alternative paths and at the same time show variations between these paths.
	An alternative for future methods, would be to specify a range of costs and compute the aggregation of the costs and superimpose these aggregations and compute the path by backtracking from the superimposed cost raster. 
	This again increases the computation time.

>>>>

	\begin{abstract}
		Least Cost Paths for a power line from a power station at the shore line to a power station further inland are calculated.
		Downsampling and clipping are investigated as techniques to speed up the calculation.
		The paths from the clipped cost raster often are the same as the original paths.
		The speed-up of the clipping can not be known in advance.
	\end{abstract}

@@

	%%
	%% Keywords. The author(s) should pick words that accurately describe
	%% the work being presented. Separate the keywords with commas.
	\keywords{GIS, Least Cost Path}

@@

	However, it is not enough to find the shortest route when building new power lines.
	Other parameters have to be taken into consideration, as the steepness of a road or the soil for example, play an important role for the building cost of a road or pipeline~\cite{suleiman_optimal_2015}.
	When planing the additional routes for a power grid, further aspects such as legal regulations and acceptance
	by the local population have to kept in mind.
	Also technical aspects, as the effects on the grid stability are further points to take into consideration~\cite{schafer_understanding_2022}.

	Due to the increasing demand for renewable energy wind offshore wind turbines supplied 5.5~\% annual percentage in 2020 of the German energy mix~\cite{noauthor_nettostromerzeugung_2021}.

	 The physical modulation in the power systems models, range from not modelling the grid at all to models  The use of Kirchhoffs laws~\cite{bertsch_participatory_2016}.
	The calculation of a power system is complex, because changing one edge changes the flow in all parts of grid.
	Besides, the actual grid data are confidential~\cite{bertsch_participatory_2016}.
	Some models not only consider the grid of one nation, but consider neighbouring states or the whole European power system~\cite{dietrich_will_2010}.	
	Older power grid models are e.g.\ optimising the well fare, but economic modelling is not sufficient for planning modern routes, as the aspect of environmental sustainability, security of supply and the public acceptance play an increasingly important role~\cite{hauff_gesellschaftliche_2011} in modern planning.

@@

	The uncertainties associated with \acrshort{MCDA} are data uncertainties, preferential uncertainties and model uncertainties are investigated with a sensitivity analysis~\cite{bertsch_participatory_2016} or simulation~\cite{butler_simulation_1997}.
	Both inter- and intra-criteria preferential uncertainties~\cite{bertsch_sensitivity_2007} can be considered.

@@

	\section{Conclusion}\label{sec:conclusion}
	The cost of computing of the Least Cost Path scales with the square of the resolution.
	The difference between the aggregated costs of the paths per resolution computed from the all~touched set to False raster True raster only shows a linear decrease.
	Therefore, the gain in accuracy per compute time decreases.
	The presented methods attempt to circumvent this bottleneck.
	When these strategies are applied to a medium resolution raster, the compute time for high resolution results can successfully be reduced, without compromising the run time for worse distances to the Least Cost Path from the high resolution raster.
	The paths can change significantly even for a small change in total cost.
	Therefore, to develop an alternative backtracking algorithm, that generates a corridor of costs might be a good strategy, to offer alternative paths and at the same time show variations between these paths.
	An alternative would be to specify a range of costs and compute the aggregation of the costs and superimpose these aggregations and compute the path by backtracking from the superimposed cost raster. 
	This again increases the computation time.
